---
title: 'Lab 4: Intro to ts wrangling & viz'
author: "Roupen Khanjian"
date: "2/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library(feasts)
library(slider)
```

## Read in data

```{r}

toolik <- read.csv(here("data", "toolikweather.csv"))

# ggplot(data = toolik, aes(x = date, y = mean_airtemp)) +
#   geom_line()
# need to chnage date to date column from char

```

### Convert this into a tsibble

```{r}

toolik_ts <- toolik %>% 
  mutate(date = mdy(date)) %>% 
  as_tsibble(key = NULL, index = date) # if had obvious grouping can include key
# idnex is whatever column has the time-series info. 

```

```{r}
ggplot(data = toolik_ts, aes(x = date, y = mean_airtemp)) +
  geom_line()
```

## Use `index_by()` function to aggregate time series info by intervals we specify 

```{r}

toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% # across entire index of ts, group by year and month
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))

ggplot(data = toolik_month, aes(x = yr_mo, y = monthly_mean_temp)) +
  geom_line() +
  facet_wrap(~month(yr_mo, label = TRUE))

```

aggregate by week:

```{r}
toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_air_temp = mean(mean_airtemp, na.rm = TRUE),
            min_air_temp = min(mean_airtemp))

```


## Filtering time series

Use `filter_index()` to specify ranges of time series to keep or exclude. 


```{r}

toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10")

```

```{r}
toolik_ts %>% 
  filter_index("2006-04-10" ~ "2006-05-15")
```

## Seaonplots

```{r}
toolik_ts %>% 
  filter(year(date) > 2014) %>% 
  gg_season(y = mean_airtemp)

toolik_month %>% 
  gg_season(y = monthly_mean_temp) + # ggseason is a wrapper for ggplot, can add any ggplot functions to it
  theme_light()

```

## Seasonal subseries plot

```{r}
toolik_month %>% 
  gg_subseries(monthly_mean_temp)
# can help pull out trends from month, thats hard to see in the raw data
```

## How to calculate moving averages with tsibbles

Using `slider` package in R

```{r}

set.seed(2021)
test <- rnorm(100, mean = 40, sd = 10)

slide(test, ~.x, .before = 2, .after = 2)

# w5 <- as.numeric(slide(test, mean, .before = 2, .after = 2))
# w5

```



```{r}
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp, 
                                   mean,
                                   .before = 7,
                                   .after = 7)))

ggplot() +
  geom_line(data = toolik_ts, aes(x = date, y = mean_airtemp), 
            size = 0.2, color = "gray") +
  geom_line(data = roll_toolik_15, aes(x = date, y = ma_15d), 
            color = "orange") +
  theme_minimal()


```

## Create the ACF

```{r}
toolik_ts %>% 
  ACF(mean_airtemp) %>% 
  autoplot() # if lages are days, here lags are very correlted with makes sense. 
 #days within a month of each other will be strongly positvely correlated

toolik_month %>% 
  ACF(monthly_mean_temp) %>% # lag in months, more informative
  autoplot()
```

## Decomposition by STL (seasonal, trend, and loess)

```{r}

toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf)))

head(toolik_dec) # not helpful

components(toolik_dec) %>% autoplot() # that block is the same on all the parts of the STL components

```







